---
title: "Exploratory Data Analysis"
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(knitr)
ames <- read.csv("data/ames.csv")
```

## Overview

Before building predictive models, we explore the data to understand distributions, identify outliers, and discover initial patterns.

---

## Dataset Summary

```{r summary_stats}
cat("=== DATASET OVERVIEW ===\n\n")
cat("Total homes:", nrow(ames), "\n")
cat("Time period:", min(ames$Yr.Sold), "-", max(ames$Yr.Sold), "\n")
cat("Variables:", ncol(ames), "\n\n")

cat("=== PRICE SUMMARY ===\n")
summary(ames$SalePrice)
```

**Key observations:**
- Dataset contains **`r nrow(ames)` homes** sold over **5 years** (2006-2010)
- Median price: **`r dollar(median(ames$SalePrice))`**
- Price range: **`r dollar(min(ames$SalePrice))` to `r dollar(max(ames$SalePrice))`**
- Mean > Median suggests right skew (some expensive homes)

---

## Response Variable: Sale Price

```{r price_distribution, fig.width=12, fig.height=5}
p1 <- ggplot(ames, aes(x = SalePrice)) +
  geom_histogram(bins = 50, fill = "#3498db", alpha = 0.8) +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Distribution of Sale Prices",
       x = "Sale Price", y = "Count") +
  theme_minimal()

p2 <- ggplot(ames, aes(x = SalePrice)) +
  geom_boxplot(fill = "#3498db", alpha = 0.8) +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Boxplot: Identifying Outliers",
       x = "Sale Price") +
  theme_minimal() +
  theme(axis.text.y = element_blank())

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**What we see:**
- **Right-skewed distribution** - most homes under $200K, long tail to $700K+
- **Outliers present** - several homes above $500K
- **Typical home:** $120K-$220K range (captures majority)

**Why this matters:** Right skew means mean is pulled up by luxury homes. Median is better measure of "typical" price.

---

## Key Continuous Variables

### Living Area (Square Footage)

```{r living_area}
ggplot(ames, aes(x = Gr.Liv.Area, y = SalePrice)) +
  geom_point(alpha = 0.4, color = "#3498db") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Price vs. Living Area",
       subtitle = "Strong positive relationship (r = 0.71)",
       x = "Above Ground Living Area (sq ft)",
       y = "Sale Price") +
  theme_minimal()

cat("Correlation:", cor(ames$Gr.Liv.Area, ames$SalePrice), "\n")
cat("Mean living area:", mean(ames$Gr.Liv.Area), "sq ft\n")
cat("Range:", min(ames$Gr.Liv.Area), "-", max(ames$Gr.Liv.Area), "sq ft\n")
```

**Interpretation:** As living area increases, price increases. The relationship is roughly linear. Correlation of 0.71 is strong.

### Overall Quality Rating

```{r quality}
ggplot(ames, aes(x = factor(Overall.Qual), y = SalePrice)) +
  geom_boxplot(fill = "#27ae60", alpha = 0.7) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Price by Quality Rating",
       subtitle = "Quality = subjective rating by assessor (1=Poor, 10=Excellent)",
       x = "Overall Quality", y = "Sale Price") +
  theme_minimal()

# Mean price by quality
quality_summary <- ames %>%
  group_by(Overall.Qual) %>%
  summarise(
    Count = n(),
    Mean_Price = mean(SalePrice),
    Median_Price = median(SalePrice)
  )

kable(quality_summary,
      col.names = c("Quality", "# Homes", "Mean Price", "Median Price"),
      caption = "Price by Quality Level")
```

**Interpretation:** Clear step-wise increase in price with quality. Quality 9-10 homes average $300K+, while quality 4-5 average $120K. This will be our strongest predictor.

---

## Key Categorical Variables

### Central Air Conditioning

```{r central_air}
ames %>%
  group_by(Central.Air) %>%
  summarise(
    Count = n(),
    Percent = paste0(round(n()/nrow(ames)*100, 1), "%"),
    Mean_Price = mean(SalePrice)
  ) %>%
  kable(col.names = c("Central Air?", "Count", "% of Total", "Mean Price"),
        caption = "Homes by Central Air Status")

ggplot(ames, aes(x = Central.Air, y = SalePrice, fill = Central.Air)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_manual(values = c("N" = "#e74c3c", "Y" = "#27ae60")) +
  labs(title = "Central Air Impact on Price",
       x = "Has Central Air?", y = "Sale Price") +
  theme_minimal()
```

**Interpretation:** 93% of homes have central AC. Those with AC sell for ~$85K more on average. However, this might be confounded with age/quality (newer homes have AC).

### Neighborhood

```{r neighborhood, fig.width=12, fig.height=8}
# Top 10 neighborhoods by count
top_neighborhoods <- ames %>%
  count(Neighborhood, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Neighborhood)

ames %>%
  filter(Neighborhood %in% top_neighborhoods) %>%
  group_by(Neighborhood) %>%
  summarise(
    Count = n(),
    Median_Price = median(SalePrice)
  ) %>%
  arrange(desc(Median_Price)) %>%
  ggplot(aes(x = reorder(Neighborhood, Median_Price), y = Median_Price)) +
  geom_col(fill = "#9b59b6", alpha = 0.8) +
  coord_flip() +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Top 10 Neighborhoods by Median Price",
       x = "Neighborhood", y = "Median Sale Price") +
  theme_minimal()
```

**Interpretation:** Significant price variation by neighborhood. NridgHt and NoRidge are premium areas ($300K+), while MeadowV and BrDale are budget areas ($100K).

---

## Missing Data Assessment

```{r missing_data}
missing_summary <- ames %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  filter(Missing_Count > 0) %>%
  mutate(Percent_Missing = round(Missing_Count / nrow(ames) * 100, 1)) %>%
  arrange(desc(Missing_Count))

if(nrow(missing_summary) > 0) {
  kable(head(missing_summary, 10),
        col.names = c("Variable", "Missing Count", "% Missing"),
        caption = "Variables with Missing Data (Top 10)")
} else {
  cat("No missing values in key variables!\n")
}
```

**Handling strategy:**
- **Pool Quality, Fence, etc.:** NA means "not applicable" (no pool/fence)
- **Garage, Basement:** NA means "no garage/basement" - treat as 0 sq ft
- **LotFrontage:** Has genuine missing values - will impute with median

---

## Outlier Investigation

```{r outliers}
# Identify extreme outliers (> $500K)
outliers <- ames %>%
  filter(SalePrice > 500000) %>%
  select(SalePrice, Gr.Liv.Area, Overall.Qual, Neighborhood, Year.Built)

cat("Found", nrow(outliers), "homes priced above $500K:\n\n")
kable(outliers,
      col.names = c("Price", "Living Area", "Quality", "Neighborhood", "Year Built"),
      caption = "Luxury Home Outliers")
```

**Decision:** Keep these outliers—they're legitimate luxury home sales, not data errors. They're larger (3,000+ sq ft), higher quality (9-10), and in premium neighborhoods.

---

## Key Insights from EDA

**1. Response Variable:**
- Right-skewed price distribution (median = `r dollar(median(ames$SalePrice))`)
- Outliers present but legitimate (luxury homes)

**2. Strongest Predictors Identified:**
- Overall Quality (clear step-wise pattern)
- Living Area (strong linear correlation)
- Neighborhood (3x price difference between best/worst)

**3. Data Quality:**
- Minimal true missing data (<5% in most variables)
- NA values often mean "not applicable" rather than missing

**4. Next Steps:**
Based on EDA findings, we proceed to:
- [Price Drivers](price-drivers.qmd) - Test hypotheses about AC, quality, age
- [Pricing Model](pricing-model.qmd) - Build predictive regression model