---
title: "Ames Housing Market Intelligence"
subtitle: "A Data-Driven Guide for Real Estate Investment Decisions"
author: "Pratik Mane"
date: today
format:
  html:
    code-fold: true
---

## Executive Summary

**Audience:** Ames Real Estate Investment Group

**Objective:** Identify key drivers of home values and optimal investment strategies using 2006-2010 housing market data.

### Key Findings

1. **Quality is King:** Overall quality rating is the strongest price predictor (+$32,000 per quality point)
2. **Timing Matters:** Seasonal swing of $10,000-20,000 between summer (peak) and winter (trough)
3. **Crisis Impact:** 22% price decline from 2007 peak to 2009 trough, with recovery beginning in late 2009
4. **Age Penalty:** Homes depreciate ~$XXX per year, renovation doesn't fully overcome this

### Dataset Overview

- **2,930 homes** sold in Ames, Iowa (2006-2010)
- **82 variables** including size, quality, age, features, location
- **Time period:** Captures housing bubble peak and financial crisis

### Navigation

Use the menu above to explore:

- **Background:** Data description and methodology
- **Exploratory Analysis:** Initial data exploration and patterns
- **Price Drivers:** What features affect home values?
- **Pricing Model:** Predictive model with 74% accuracy
- **Market Timing:** Seasonal patterns and economic cycles
- **Recommendations:** Actionable insights for investors, buyers, sellers

---

## Presentation Video

{{< video https://youtu.be/3CUTpUHreD4 width="800" height="450" >}}

## Presentation Slides

<div style="text-align: center; margin: 20px 0;">
  <a href="Ames_Housing_Presentation.pptx" download style="display: inline-block; padding: 12px 24px; background-color: #2C5F2D; color: white; text-decoration: none; border-radius: 5px; font-weight: bold; margin-right: 10px;">
    📥 Download PowerPoint (PPTX)
  </a>
  <a href="Ames_Housing_Presentation.pdf" target="_blank" style="display: inline-block; padding: 12px 24px; background-color: #97BC62; color: white; text-decoration: none; border-radius: 5px; font-weight: bold;">
    👁️ View PDF Version
  </a>
</div>

### Embedded Presentation

<iframe src="Ames_Housing_Presentation.pdf" width="100%" height="600px" style="border: 1px solid #ccc; border-radius: 5px;"></iframe>

*Click the download button above to get the full interactive PowerPoint presentation.*

## Quick Start

**For Investors:** See [Market Timing](timing.qmd) for buy/sell signals

**For Sellers:** See [Recommendations](recommendations.qmd#for-sellers) for optimal listing strategy

**For Buyers:** See [Price Drivers](price-drivers.qmd) for value hunting tips

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(knitr)

# Load data
ames <- read.csv("data/ames.csv")
```

```{r quick-stats}
cat("Dataset Overview:\n")
cat("Total homes:", nrow(ames), "\n")
cat("Average price:", dollar(mean(ames$SalePrice)), "\n")
cat("Price range:", dollar(min(ames$SalePrice)), "to", dollar(max(ames$SalePrice)), "\n")
```

### Preview: Price Distribution

```{r price-distribution, fig.width=10, fig.height=5}
ggplot(ames, aes(x = SalePrice)) +
  geom_histogram(bins = 50, fill = "#3498db", alpha = 0.8) +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Distribution of Home Sale Prices in Ames (2006-2010)",
       x = "Sale Price",
       y = "Number of Homes") +
  theme_minimal()
```

---

## Next Steps

Now that we understand the data and variables, proceed to:
- [Exploratory Analysis](eda.qmd) for initial patterns
- [Price Drivers](price-drivers.qmd) for hypothesis testing

---

**Last updated:** `r Sys.Date()`