---
title: "What Drives Home Prices?"
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(knitr)
library(car)
ames <- read.csv("data/ames.csv")
```

## Overview

This analysis identifies which home features have the strongest impact on sale prices using:
- **Correlation analysis** (Week 6)
- **Hypothesis testing** (Week 7)
- **ANOVA** (Week 8)

---

## Top Price Drivers (Correlation Analysis)

```{r correlations}
# Calculate key correlations
correlations <- data.frame(
  Feature = c("Overall Quality", "Living Area", "Garage Area", 
              "Basement Area", "Year Built"),
  Correlation = c(
    cor(ames$Overall.Qual, ames$SalePrice),
    cor(ames$Gr.Liv.Area, ames$SalePrice),
    cor(ames$Garage.Area, ames$SalePrice, use = "complete.obs"),
    cor(ames$Total.Bsmt.SF, ames$SalePrice, use = "complete.obs"),
    cor(ames$Year.Built, ames$SalePrice)
  )
) %>%
  arrange(desc(abs(Correlation)))

kable(correlations, digits = 3,
      caption = "Top 5 Features Correlated with Sale Price")
```

**Key Finding:** Overall Quality (r = 0.80) and Living Area (r = 0.71) are the strongest predictors.

### Visualization: Quality vs. Price

```{r quality-price, fig.width=10, fig.height=6}
ggplot(ames, aes(x = factor(Overall.Qual), y = SalePrice)) +
  geom_boxplot(fill = "#3498db", alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23, 
               size = 3, fill = "red") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Sale Price by Overall Quality Rating",
       subtitle = "Red diamond = mean price for each quality level",
       x = "Overall Quality (1 = Poor, 10 = Excellent)",
       y = "Sale Price") +
  theme_minimal()
```

---

## Feature Impact: Hypothesis Tests

### Central Air Conditioning

```{r central_air}
# Create the groups FIRST
ac_yes <- ames %>% filter(Central.Air == "Y") %>% pull(SalePrice)
ac_no <- ames %>% filter(Central.Air == "N") %>% pull(SalePrice)

# Then run the t-test
t_result <- t.test(ac_yes, ac_no)

cat("Homes WITH Central Air: mean =", dollar(mean(ac_yes)), "\n")
cat("Homes WITHOUT Central Air: mean =", dollar(mean(ac_no)), "\n")
cat("Difference:", dollar(mean(ac_yes) - mean(ac_no)), "\n")
cat("P-value:", format(t_result$p.value, scientific = TRUE), "\n")
```

**P-value interpretation:**

The p-value of `r format(t_result$p.value, scientific = TRUE)` means:

- **If central air had NO effect** on price (null hypothesis true)
- The probability of observing a difference of `r dollar(round(mean(ac_yes) - mean(ac_no)))` or more extreme
- **Is less than 0.001** (0.1%)

In other words: This result is so extreme that it's virtually impossible to have occurred by chance alone if AC truly didn't matter. We have **very strong evidence** that central air increases home values.

**Statistical significance:** ✓ p < 0.001 (highly significant)  
**Practical significance:** `r dollar(round(mean(ac_yes) - mean(ac_no)))` difference is substantial (approximately 47% of median home price)

---

### Recent Construction (2000+)

```{r recent_construction}
# Create binary variable for recent construction
ames <- ames %>%
  mutate(Is_Recent = Year.Built >= 2000)

recent <- ames %>% filter(Is_Recent) %>% pull(SalePrice)
older <- ames %>% filter(!Is_Recent) %>% pull(SalePrice)

t_result2 <- t.test(recent, older, alternative = "greater")

cat("Recent homes (2000+): mean =", dollar(mean(recent)), "\n")
cat("Older homes (<2000): mean =", dollar(mean(older)), "\n")
cat("Difference:", dollar(mean(recent) - mean(older)), "\n")
cat("P-value:", format(t_result2$p.value, scientific = TRUE), "\n")
```

**P-value interpretation:**

The p-value of `r format(t_result2$p.value, scientific = TRUE)` means:

- **If construction year had NO effect** on price (null hypothesis true)
- The probability of observing a difference of `r dollar(round(mean(recent) - mean(older)))` or more extreme
- **Is less than 0.001** (0.1%)

**Finding:** Recently built homes (2000+) sell for **`r dollar(round(mean(recent) - mean(older)))`** more on average than older homes.

**Statistical significance:** ✓ p < 0.001 (highly significant)  
**Practical significance:** This $92K+ premium represents a 51% increase over the average older home price

---

## House Style Effect (ANOVA)

```{r anova_style}
# ANOVA test
anova_result <- aov(SalePrice ~ House.Style, data = ames)
anova_summary <- summary(anova_result)

cat("ANOVA F-statistic:", round(anova_summary[[1]]["House.Style", "F value"], 2), "\n")
cat("P-value:", format(anova_summary[[1]]["House.Style", "Pr(>F)"], scientific = TRUE), "\n")
cat("\n✓ House style DOES affect price (p < 0.001)\n")

# Mean price by style
style_summary <- ames %>%
  group_by(House.Style) %>%
  summarise(
    Count = n(),
    Mean_Price = mean(SalePrice),
    .groups = 'drop'
  ) %>%
  arrange(desc(Mean_Price))

kable(style_summary, 
      col.names = c("House Style", "Count", "Mean Price"),
      digits = 0)
```

**P-value interpretation:**

The p-value < 0.001 from the ANOVA test means:

- **If house style had NO effect** on price (all styles have same mean price)
- The probability of observing differences this large between groups
- **Is less than 0.1%**

**Finding:** Two-story homes command highest prices (mean: ~$200K), while split-level/multi-level homes have lowest prices (mean: ~$140K).

**Statistical significance:** ✓ p < 0.001 (highly significant)  
**Practical significance:** $60K+ difference between highest and lowest style types

---

## Summary: Price Driver Rankings

**Top 5 drivers (by effect size):**

1. **Overall Quality** (r = 0.80) - Strongest predictor
2. **Living Area** (r = 0.71) - Size matters
3. **Recent Construction** (+$92K) - Buyers pay premium for new
4. **Central Air** (+$85K) - Modern necessity
5. **Garage Area** (r = 0.64) - Storage/parking valued

**Actionable insight:** Focus on quality improvements over size additions for best ROI.

---

## Next Steps

Proceed to [Pricing Model](pricing-model.qmd) to build a multiple regression model that controls for these features simultaneously and quantifies their independent effects.
