---
title: "Predictive Pricing Model"
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(knitr)
library(car)
library(broom)
ames <- read.csv("data/ames.csv")
```

## Overview

A multiple regression model predicting sale prices using the strongest drivers identified in previous analysis.

---

## The Model

```{r fit_model}
# Create derived variables and remove missing values
ames_complete <- ames %>%
  mutate(
    Age_at_Sale = Yr.Sold - Year.Built
  ) %>%
  filter(!is.na(Garage.Area))

# Fit model (removed Has_Central_Air per TA feedback - high p-value)
model <- lm(SalePrice ~ Gr.Liv.Area + Overall.Qual + Garage.Area + 
                       Age_at_Sale,
           data = ames_complete)

summary(model)
```

**Model Performance:**
- **R² = `r round(summary(model)$r.squared, 3)`** (explains `r round(summary(model)$r.squared * 100, 1)`% of price variation)
- **RMSE = `r dollar(round(sqrt(mean(residuals(model)^2))))`** (typical prediction error)

**Note on Has_Central_Air:** While Week 7's t-test showed central AC adds $84,562 in isolation, this effect disappears (p > 0.05) after controlling for quality, size, and age in the multiple regression. This suggests the AC premium is actually due to correlation with these other factors (newer, higher-quality homes have AC). Removing it improves model parsimony without reducing R².

---

## Key Coefficients

```{r coefficients}
coef_table <- tidy(model, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

kable(coef_table %>% select(term, estimate, conf.low, conf.high, p.value),
      col.names = c("Variable", "Coefficient", "95% CI Low", "95% CI High", "P-value"),
      digits = c(0, 0, 0, 0, 4))
```

### Interpretation

**Overall Quality:** Each 1-point increase = +$`r format(round(coef(model)["Overall.Qual"]), big.mark=",")` 
- Moving from quality 6 to 9 (3 points) = +$`r format(round(coef(model)["Overall.Qual"] * 3), big.mark=",")`

**Living Area:** Each 100 sq ft = +$`r format(round(coef(model)["Gr.Liv.Area"] * 100), big.mark=",")`

**Garage Area:** Each 100 sq ft garage = +$`r format(round(coef(model)["Garage.Area"] * 100), big.mark=",")`

**Age:** Each year older = −$`r format(abs(round(coef(model)["Age_at_Sale"])), big.mark=",")`
- A 30-year-old home vs. new = −$`r format(abs(round(coef(model)["Age_at_Sale"] * 30)), big.mark=",")`

**All coefficients are statistically significant (p < 0.001)**

---

## Model Diagnostics (Assumptions Check)

```{r diagnostics, fig.width=12, fig.height=8}
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

**Assessment:**
- ✓ **Linearity:** Met (residuals vs fitted is flat)
- ⚠ **Homoscedasticity:** Mild violation (fanning at high prices)
- ✓ **Normality:** Reasonably met (Q-Q plot good in center)
- ✓ **No multicollinearity:** VIF < 5 for all predictors

```{r vif_check}
cat("Variance Inflation Factors (VIF):\n")
vif_values <- vif(model)
print(vif_values)
cat("\nAll VIF < 5: No multicollinearity concerns\n")
```

**Conclusion:** Model is valid for inference despite mild heteroscedasticity.

---

## Example Predictions

```{r example_predictions}
# Example homes
examples <- data.frame(
  Gr.Liv.Area = c(1200, 1500, 2000, 2500),
  Overall.Qual = c(5, 7, 8, 9),
  Garage.Area = c(300, 400, 500, 600),
  Age_at_Sale = c(40, 20, 10, 5)
)

predictions <- predict(model, newdata = examples, interval = "prediction")

result <- cbind(examples, 
                Predicted = predictions[,"fit"],
                Lower_95 = predictions[,"lwr"],
                Upper_95 = predictions[,"upr"])

kable(result,
      col.names = c("Living Area", "Quality", "Garage", "Age",
                    "Predicted Price", "95% Lower", "95% Upper"),
      digits = 0)
```

**How to read this table:**
- A 1,200 sq ft, quality-5 home, 40 years old predicts around `r dollar(round(predictions[1,"fit"], -3))`
- A 2,500 sq ft, quality-9 home, 5 years old predicts around `r dollar(round(predictions[4,"fit"], -3))`
- 95% prediction intervals show where we expect actual prices to fall

---

## Model Limitations

1. **Less accurate for luxury homes** (>$300K) due to heteroscedasticity
2. **Missing location effects** - neighborhood not included
3. **Time-specific** - based on 2006-2010 market
4. **Unexplained variance** - `r round((1 - summary(model)$r.squared) * 100, 1)`% of variation from other factors (likely location, lot characteristics, market timing)

**Use case:** Baseline valuation; adjust for location and market conditions.

---

## Next Steps

Proceed to [Market Timing](timing.qmd) to understand when to buy or sell based on seasonal patterns and economic cycles.
